Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-02-28 Alife16/paper.tex
Reduced to 11 pages a a bit
But still a way to go. Almost everything must go refs #99

\title{ Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources 
for Low Cost Distributed Evolutionary Computation}
\author{Juan-J.~Merelo*$^1$, Mario Garc\'ia-Valdez$^2$\\
\mbox{}\\
$^1$Dept. of Computer Architecture and Technology and CITIC University of Granada \\
$^2$Dept. of Graduate Studies at Instituto Tecnol\'ogico de Tijuana \\
{\tt jmerelo@ugr.es}, {\tt mario@tectijuana.edu.mx}}

@@

\begin{abstract}
Volunteer computing is extensively used 
in science and industry, but it
still has a long way to go in order to become
 a mainstream
technology in the same way the cloud technologies in general have. The
lack of 
general purpose and open source tools with built-in analytics 
could be one of the reasons of why this is so, as well as the lack of
proofs of concept. That is why in this paper we present {\sf NodIO}, a
framework used to create client-server architectures for distributed
evolutionary algorithms, entirely written  
in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser, as well as any other client consuming its REST application
programming interface. An enhanced implementation, {\sf NodIO-W$^2$} is also
presented, this version leverages the HTML 5 Web Workers specification
to execute several clients per browser in the background, so that the the number of
generations per user could be improved. The architecture, along with some initial
results on the expected performance and scaling behavior when the
number of users is increased is presented. We also use
the algorithm data to find out what kind of patterns the user behavior
follows by examining the resources they devote to our experiment, and
try to fit them to some statistical distribution to create the bases
of a model of user behavior in volunteer computing experiments of this
kind that can be used by researchers to predict its performance. 
In general, our experiments prove, first, that {\sf NodIO} is a flexible and
low-overhead platform to perform distributed computation experiments
that include volunteers, and then that user behavior follows some patterns,
and, thus, are amenable to a model.
\end{abstract}

@@

\section{{\sf NodIO-W$^2$}, a Web Worker-based architecture}
\label{sec:w2}

In order to improve the number of cycles per user, we enhanced the
basic architecture in several different aspects.  One improvement 
involved restarting the client once the solution was
found, so that as long as the user was visiting the page it was
running an evolutionary algorithm. Another was a small algorithmic
enhancement: instead of having a fixed population,
population size was randomly distributed between 128 and 256 and
different for each client.  The other improvement was to 
use the Web Worker API as described next.

JavaScript has been commonly used to develop client-side scripts
that handle the user interface, communicating asynchronously with servers and
updating the content that is displayed \citep{flanagan2006javascript}.
Long-running scripts like those proposed in this work can interfere with the
responsiveness of the web application, because everything is handled by a
single thread. The Web Workers specification \citep{hickson2012web} defines an
API that allows Web developers to execute long-running scripts in parallel
with the scripts of the main page. Worker scripts are spawned in the
background allowing a thread-like operation using message-passing as the
communication mechanism between the worker's execution environment and the
main thread. According to the Web Workers W3C Candidate Recommendation
\citep{hickson2012web} workers are expected to be long-lived, they have a high
start-up performance cost, and a high per-instance memory cost. The Web Worker
API is implemented in current versions of both desktop and mobile web browsers.
For volunteer computing using the Web Worker API brings several advantages
over a single threaded implementation:

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even in case the user brings forward another tabbed
window.
\item Several evolutionary algorithms can be executed in parallel in a single web
page. 
\item Many implementations of the Web Worker API can use multiple-core CPUs for
the execution of worker scripts, in particular the execution in multiple-core CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each {\sf NodEO} instance can be restarted independently.
\end{itemize}

A sequence diagram of the enhanced NodIO, which is called NodIO-W$^2$, is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two {\sf NodEO} instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this system there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread locally in the client, which are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em Node.js} This is the server side process which is basically
  the same as used previously, except for some minor log-related
  changes. 
\item {\em Client main thread} The main script creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the
XMLHttpRequest object.  
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=3in]{Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}
%
\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments is initialized and a log
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP
request, then the page is loaded.
\item The main script renders the page and creates the worker instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations the state of the population
is evaluated. The chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared population. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the target. If the best fitness is a global best, the best
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by displaying a dynamic plot of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The red dot and line show the average and standard
  deviation, the shade of blue the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
\end{figure}

The experiments were run in the same way as before, using several
Twitter announcements throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most 
of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table*}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of {\tt PUT}s per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%

However, it is interesting to note why this is so, and the first hint
is the inter-experiment correlation between the number of IPs in
successive experiments. While before it was an unremarkable 10\%, it is
now more than 43\%, making the number of IPs in contiguous experiments
highly correlated. The main reason for this is the setup in {\sf
  NodIO-W$^2$}, which restarts an island after a solution has been
found and also that it keeps running even if the tab is not on the
foreground. This means that volunteers can leave the experiment
running for as long as they want and they will be contributing 
experiment after experiment, unlike the previous version, where they
stopped contributing after one solution was found. This is also
reflected in the number of PUTs, which is the number of generations
divided by 100; more than 50\% of the volunteers run the experiment for more than
4000 generations. All things considered, this means than there are
many more computers {\em simultaneously} running, leading to this
almost 5-fold increase in running time.

We are also interested in measuring the scaling properties in this
model, that is, the relationship between the number of IPs
participating in an experiment and the time it takes to complete
it. This is shown in Figure \ref{fig:ipstime:w2}, which displays every
experiment in terms of time (in milliseconds) vs. the number of workers participating
in it. Although there is not a clear trend, the graph seems to say
that the time needed for finding the solution in this evolutionary
algorithm does not depend on the number of workers participating in
it. This does not mean that it is independent on the number of {\em
  simultaneous} users, which is probably the case although it is more
difficult to measure. There seems to be a trend towards a decreasing
standard deviation in the time, with more volunteers adding both {\em
  robustness} to an experiment, and certainty in the time it is taking. 
 However, more experiments would be needed to check this hypothesis.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of {\tt PUT}s per worker and GEV fit (in red, right) 
and Weibull (in red, left) for the {\sf NodIO-W$^2$} model.}  
\label{fig:gev:w2}
% File ../data/ips-puts.R
\end{figure}
%
As we did in the previous subsection, we have also fitted the number
of HTTP {\tt PUT}s per worker to a  Generalized Extreme Value (GEV) function, with the result shown
in Table \ref{tab:puts:ww}. Comparing this table with what we obtained
in the previous experiment, Table \ref{tab:puts:os}, we find higher values
for the location $\mu$ and scale $\sigma$ parameters, accounting for
the higher number of contributions per volunteer that has been
previously observed. However, the shape $\xi$ parameter is
substantially similar, with a {\em bell} shape leaning towards the
origin, indicating a pattern of a few users with many contributions
and many with few contributions. If we plot the model and the data
side by side, as shown in Figure \ref{fig:gev:w2} we see that the fit
is not so tight as in the previous case, with the model overestimating
the number of individuals with a high number contributions. This is
probably due to the fact that in the previous case, the user needed to
voluntarily reload the page to contribute the most generations, with a
{\em phase change} between the people that did so and the people that
just ran the experiment once until completion. In this case, however, the
user just needs to let it run, without that phase change and thus we obtain a
more {\em egalitarian} distribution of contributions, which rather
corresponds to a Weibull function, as shown in Figure
\ref{fig:ipstime:w2} (right hand side). The fitted parameters for this
distribution, shown in Table \ref{tab:puts:ww}, show a shape parameter
equal to 0.69, a value that is remarkably similar to the 0.5 value
found for time devoted to games in Chambers et
al. \citep{chambers2005measurement}. It is important to note that, in
this case, the number of PUTs made per client does not correspond to a
uniform amount of computation, since 100 generations with a random
population belong, in each case, to different number of
operations. In other words, the same number of PUTs will correspond to
different number of operations in each case. However, since clients
are also different and take a different time in each case, it is
difficult to ascertain how this might have an influence in the
statistical distribution.

In general, this second version of the {\sf NodIO} framework and the
single experiment performed prove that it can be the foundation for a distributed
high-server-performance evolutionary computation platform,
providing reasonable algorithmic performance and being, in general,
easy to use with a straightforward modification of the fitness
function.

\subsection{$F_{15}:\frac{D}{m}$-group Shifted and $m$-rotated Rastrigin's Function}
JavaScript has been traditionally implemented as an interpreted
language not designed for the development of high performance
systems. On the other hand, current JavaScript  
virtual machines (VMs) are closing the gap, for instance the Google V8 
is an engine specifically designed for the fast execution of 
large JavaScript applications. In order to increase performance
V8 compiles JavaScript source code directly into machine code when it is first executed. 
There are no intermediate byte codes or an interpreter \citep{Gray:2009:GCM:1610564.1610565}.
JavaScript has gained popularity in server-side development using runtime
environments such as Node.js and is also used in desktop and mobile applications. 
%Mario: Creo que el párrafo de arriba sale sobrando
% (Paloma) Agree, I mean, it adds information but in case you have space problems... is ok if you start the section without it
To further evaluate how a current JavaScript implementation could be used by a
computer scientist to develop complex optimization problems,
a benchmark function, which was provided by the CEC2010 Special Session on
Large-Scale Global Optimization \citep{tang2007benchmark} is used in this section. 
The function is described next.

The basic Rastrigin's function is separable, and is defined as follows;
\begin{equation}
F_{rastrigin}(x)=\sum\limits_{i=1}^D [ x_{i}^{2}-10\cos(2\pi x_i)+10  ] 
\end{equation}
where $D$ is the dimension and $x = (x1, x2, · · · , x_{D})$ is a
$D$-dimensional row vector (i.e., a $1 × D$ matrix). Rastrigin’s function 
is a classical multimodal problem. Such problem is difficult since the number of local
optima grows exponentially with the increase of dimensionality. To make it 
non-separable, an orthogonal matrix is also used for coordinate rotation.
The rotated Rastrigin's function is defined as follows:
\begin{equation}
F_{rot\_rastrigin}(x)=F_{rastrigin}(\textbf{z}), \textbf{z}= \textbf{x} * \textbf{M}
\end{equation}
Where $\textbf{M}$ is  a $D \times{D}$ orthogonal matrix. The benchmark
emulates real-world optimization problems that most likely will consist 
of different groups of parameters with strong dependencies within
but little interaction between the groups. This issue is reflected in the benchmark 
by randomly dividing the objective variables
into several groups, each of which contains a number of variables. The parameter $m$ 
is used to control the number of variables in each group and hence, defining the degree
of separability:
\begin{equation}
F_{15}(x)=\sum\limits_{k=1}^\frac{D}{m}F_{rot\_rastrigin}[\textbf{z}(P_{(k-1)*m+1}:P_{k*m})] 
\end{equation}
For the benchmark $D = 1000$, Group size $m = 50$, $\textbf{x} = (x_1, x_2, \ldots , x_{D})$ is the
candidate solution, $\textbf{o} = (o_1, o_2, \ldots  , o_{D})$ is the shifted global optimum,
$\textbf{z} = \textbf{x} - \textbf{o}$ is the shifted candidate solution and
finally $\textbf{P}$ is a random permutation of $\lbrace1, 2, \ldots  ,D\rbrace$.

This optimization problem was selected because it is representative of the kind of
algorithms, data types, and structures employed in large scale optimization problems. 
The function was implemented in both Matlab and Java languages as part of the
Test Suite of the companion competition of the Special Session and to give competitors an idea of the
computational cost of the challenge, the runtime required for 10,000 function evaluations 
for a particular configuration was published by organizers. The Java
implementation took 7596ms and the Matlab version 1115ms. The whole experiment had the value of
3,000,000 function evaluations as termination condition. Using the Java implementation as
a guide, the function described above was implemented to compare the performance of JavaScript against
both implementations, again using the time required for 10,000 function evaluations. 
Before discussing the setup and performance results, certain details of
the implementation in JavaScript are presented to give readers an idea of 
how practical and mature this language is for developing complex mathematical functions.

\begin{itemize}
\item {\em Randomize} Both Matlab and Java programs rely on a 
Java Randomizer library for the 
randomization of the shift vectors and matrices used by the function.
The library includes the generation of pseudorandom Gaussians of type
double and uniformly distributed integers. Generators use a single
{\tt long} seed. In order to implement the same functionality in
JavaScript, the library {\tt random-js} was used, as there are inconsistencies 
in the implementation of the standard {\tt Math.random()} between engines
and also their results are non-deterministic. 
The library {\tt random-js}, implements a Mersenne Twister algorithm producing 
consistent results across all JavaScript VMs.
\item {\em Timing functions} In order to compare the performance of 
each implementation, an accurate and consistent way of measuring 
the runtime of functions is needed.
In JavaScript the {\tt Date class} is often used 
to measure execution time, but has the limitation of having a maximum 
resolution given in milliseconds.
In order to measure the runtime intervals with a higher precision two functions
where used: for Node.js we used the native {\tt process.hrtime()} function
which returns a high-resolution measure in a {\tt [seconds, nanoseconds]}
array, and for measuring in the browser the {\tt Performance.now()} function was called,
which uses floating-point numbers with up to microsecond precision. 
A drawback found is that the {\tt Performance.now()} function is only implemented
in Firefox and Chrome browsers. Both functions are independent of the system clock. 
\item {\em Tools} An important consideration when choosing certain 
language to write non-trivial programs is the availability of development tools. While developing
the experiments, we used a comprehensive set of developer tools available for both
browsers and desktop \citep{tilkov2010node}. These tools included
package managers, in this case {\tt npm}, debuggers, logging libraries
such as {\tt winston}, 
network monitors, editors, and even debugging of the multi-threaded execution of Web Workers.
Although the fact that developer tools vary from browser to browser
supposed a drawback, they all have the same functionality. The
JavaScript debugger can be accessed directly from the user menu in
most browsers. 
\item {\em Data Types} JavaScript uses floating point numbers with a 
limited precision of 64 bits that partially implement the functionality of
the {\tt StrictMath} library used by the Java implementation. If more precision
is needed, developers could use {\tt math.js}, an extensive math library with
support for matrices and big numbers.
\end{itemize}

We found that the JavaScript ecosystem is mature enough for developing non-trivial 
code, but developers have to consider the differences between VMs implementations
and therefore use certain libraries in order to increase the precision and repeatability of
algorithms. As for the performance of JavaScript against Java and Matlab
we conducted first the basic test described earlier in a 3.7 GHz Quad-Core
Intel Xeon E5 processor running OS X 10.10.5, 
using Java(TM) SE Runtime Environment (build 1.8.0\_25-b17), 
MatLab version R2015a for Mac OSX, Node FrameWork version 0.12.2,
Google Chrome Version 46.0.2490.86 (64-bit) 
and Firefox version 41.0.2. % (Paloma) Table here? Or at least the characteristics could be represented the same way as for the baseline experiment
 The performance of the time required for 10,000 function evaluations is presented in
Figure \ref{fig:f15_times}. In our test the Matlab implementation had the best
performance with an average of 935ms, followed by Java with
991ms. On the other hand, JavaScript took 32\% more time than Java with 1238ms in Chrome using a 
single Web Worker and 1,234ms using Node.js. % Are these results correct? JS=>1240ms and Node.js=>1,270ms ???
The results also show that there is not much difference between the Node.js
implementation an that of a
browser. There is also not much difference between running the code in the main thread or in Web
Workers. When running the experiment in Firefox a dialog alerted that the
script was taking too much time, showing that this
kind of processing is not intended for the main thread. Additionally, there is not much overhead
when running two
Web Workers in parallel as they took 1279ms each. This experiment
shows that JavaScript is viable as a language
language for developing complex mathematical functions an these can run in a browser
with acceptable performance. 
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{f15_times.png}
\caption{ Runtime of 10,000 function evaluations for $F_{15}$} 
\label{fig:f15_times}
\end{figure}

\subsubsection{Setup of Rastrigin's experiment in {\sf NodIO-W$^2$ }}
This experiment was deployed in a virtual private server with 512MB of memory
running Ubuntu OS version 14.04 and hosted in DigitalOcean.com. Unlike previous
experiments where the optimum was expected to be found many times, 
this problem was expected to run constantly for many days finding only
sub-optimal approximations. In order to achieve this, the only change needed 
in the system was to limit the number of chromosomes kept in the pool. 
Previously, every PUT request which included a new chromosome increased 
the size of the pool, in a long-running experiment this has to be avoided
in order to not exhaust the available memory. In this experiment the size
of a chromosome was 1000 double precision numbers, considering this
and the memory available in the server, the the pool was limited 
to 10,000 chromosomes.  In order to keep this size
constant a random chromosome was pulled out of the pool before inserting a
new one, when the pool was full. For this problem {\sf NodEO}'s {\tt chromosome-float} 
and {\tt classic-ea-float} modules were used. The parameters for the EA 
were tournament size = 3 and  population size = 500, returning exchanging a
chromosome with the server every 100 generations. Again two workers in each window.
As before, a few participation calls were published in social networks.

\subsubsection{Results}

%
\begin{figure}[!htb]
\centering
\includegraphics[width=3in]{rastrigin-fitness.png}
\caption{Fitness of chromosomes sent to the server pool along time in
  the $x$ axis.} 
% Plotted with ../data/plot-rastrigin-fitness.R
\label{fig:puts:rastrigin}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{rastrigin-IPs.png}
\includegraphics[width=0.49\linewidth]{rastrigin-IPs-hour.png}
\caption{Number of unique IPs participating in the Rastrigin
  experiment per minute (left) and per hour (right)} 
% Plotted with ../data/plot-rastrigin-IPs.R
\label{fig:ips:rastrigin}
\end{figure}
%
\begin{table*}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of {\tt PUT}s per worker for the F15 (Rastrigin) function. \label{tab:puts:ww:f15}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 35.759  &  343.336   & 9.877 \\
Weibull & ND & 29.002625061 $\pm$ 2.153034198  & 0.402393903 $\pm$ 0.007586468 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww-rastrigin-workers.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww-rastrigin-workers.png}
\caption{Fit to GEV (left) and Weibull (right) of the number of {\tt PUT}s per worker for the F15 (Rastrigin) function.} 
% Plotted with ../data/ips-puts-rastrigin.R
\label{fig:fit:rastrigin}
\end{figure}
%

The results for this new experiment have been extremely interesting
and quite different from the previous one. The most important
characteristic that set this experiment apart from the previous one is
the fact that it did not find the solution for the time it was
running, instead lowering, little by little, the fitness until it was
disconnected. The fitness of chromosomes sent by clients is shown in
Figure \ref{fig:puts:rastrigin}, that shows two different parts: the
upper part, which represents the fitness of clients that have just
started evolution from a random population, and then the fitness of
those clients that have been already evolving for a while or obtained
a high-fitness individual from the pool and evolved from there, and
which managed to surpass the 9000 value by the end of the
simulation. It incidentally shows also the number of {\em new} nodes
coming: there is a dense {\em cloud} at the beginning, but the new
contributions keep thinning as the evolution progresses.

This is in part due to the fact that the presence of nodes is not
constant, as shown in Figure \ref{fig:ips:rastrigin}, which plots the
number of unique IPs participating in the experiment per minute (left)
and hour (right). The two graphs have the same shape which matches
also the speed in which fitness is decreased in Figure
\ref{fig:puts:rastrigin}. At the beginning there are very few or only
one, but later one there are peaks of almost 30 unique machine a
minute; these peaks appear again when new announcements are made in
Twitter or any other social networks, to go down to a few remaining
nodes later on. It is interesting to note that, at the most, more than
100 unique nodes were participating in the experiment during one hour;
also that there are sustained peaks during a day of 6-7 computers
every minute. This is due to the fact that the computation is
asynchronous, but also to the fact that different computers stay in
the simulation for a different period of time. This period has been
plotted in Figure \ref{fig:fit:rastrigin}, that plots the number of
{\tt PUT}s per node, ranked by number, along with a fit to the two
distributions we have used before, GEV and Weibull; the parameters of
this fit are shown in Table \ref{tab:puts:ww:f15}. Unlike the previous
experiments, in which GEV fit was relatively good, in this case it is
quite obviously not. However, the fit to the Weibull distribution,
shown at the left of the image, is relatively good. As should be
expected, the shape and scale are different to the one in the previous
experiments with the Trap function, however, the shape is quite
similar (0.40 vs 0.69), being in both cases less than one, which
indicates a decreasing distribution of the same kind. In fact, the
difference in the scale parameter indicates that the difference
between the time devoted by a volunteer and the next is bigger in this
case. This is probably due to the fact that the duration of the
experiment is bigger, allowing for bigger differences in
behavior. While the first in the Trap function might have devoted a
few minutes and the second a few seconds less, the difference between
the first and second in this experiment might be in the scale of
hours.

In general, this experiment confirms what was found in the others in
several aspects: first, there is a regularity in the distribution of the
time assigned by volunteers to it, second, there is a good and certain
amount of volunteers that will become part of the distributed
computer, in some cases up to one hundred; also, in some cases there
will be users that will spend several hours, two orders of magnitude
more generations in this case than in the previous one. There is a
clear influence of the time to solution: a short time means that every
experiment will only gather a few volunteers, but also that they will
stay only for a short time. A long time will make users stay for
longer, but if it takes too long, they will eventually lose interest.

We will discuss this and the rest of the findings of the paper in the
conclusions.

@@

In the second implementation, {\sf NodIO-W$^2$}, Web Workers
were used so that several clients per browser could be executed and
they could run in the background, so that the the number of
generations per user could be improved. The number of generations per
user increased to the point that baseline performance was improved in
most cases. The architectural changes led to a high correlation
between successive experiments, so that, even as the median number of
volunteers per experiment decreased, they were probably contributing
to the experiment at the same time, so that our objective was
achieved. In general, one of the first conclusions for this paper is
that the browser technologies should be used to its full extent so
that the user time is leveraged to its full potential, and this was
achieved with {\sf NodIO-W$^2$}.

@@

This work has been supported in part by TIN2011-28627-C04-02 and
TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
SPIP2014-01437 (Direcci{\'o}n General de Tr{\'a}fico) and PYR-2014-17
GENIL project (CEI-BIOTIC Granada). We would also like to thank the
anonymous reviewers of previous versions of this paper who have really
helped us to improve 
this paper (and our work) with their suggestions. We are also grateful
to Anna S\'aez de Tejada for her help with the data processing scripts.

>>>>

\title{ Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources 
for Low Cost Distributed Evolutionary Computation}
\author{Juan-J.~Merelo*$^1$, Paloma de las Cuevas $^1$, Pablo
  García-Sánchez $^1$, Mario Garc\'ia-Valdez$^2$\\
\mbox{}\\
$^1$Dept. of Computer Architecture and Technology and CITIC University of Granada \\
$^2$Dept. of Graduate Studies at Instituto Tecnol\'ogico de Tijuana \\
{\tt jmerelo@ugr.es}, {\tt mario@tectijuana.edu.mx}}

@@

\begin{abstract}
Volunteer computing is extensively used 
in science and industry, but it
still has a long way to go in order to become
 a mainstream
technology in the same way the cloud technologies in general have. The
lack of 
general purpose and open source tools with built-in analytics 
could be one of the reasons of why this is so, as well as the lack of
proofs of concept. That is why in this paper we present {\sf NodIO}, a
framework used to create client-server architectures for distributed
evolutionary algorithms, entirely written  
in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser, as well as any other client consuming its REST application
programming interface. The architecture, along with some initial
results on the expected performance and scaling behavior when the
number of users is increased is presented. We also use
the algorithm data to find out what kind of patterns the user behavior
follows by examining the resources they devote to our experiment, and
try to fit them to some statistical distribution to create the bases
of a model of user behavior in volunteer computing experiments of this
kind that can be used by researchers to predict its performance. 
In general, our experiments prove, first, that {\sf NodIO} is a flexible and
low-overhead platform to perform distributed computation experiments
that include volunteers, and then that user behavior follows some patterns,
and, thus, are amenable to a model.
\end{abstract}

@@

@@

@@

This work has been supported in part by TIN2011-28627-C04-02 and
TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
SPIP2014-01437 (Direcci{\'o}n General de Tr{\'a}fico) and PYR-2014-17
GENIL project (CEI-BIOTIC Granada). We would also like to thank the
anonymous reviewers of previous versions of this paper who have really
helped us to improve 
this paper (and our work) with their suggestions. We are also grateful
to Anna S\'aez de Tejada for her help with the data processing
scripts. We are also grateful to {\tt @otisdriftwood} for his help
gathering users for the new experiments.
